Microlensing as a possible probe of the internal structure of quasars

### 2015-07-12 paper1/code.tex

\section{Microlensing code and magnification maps}

We now go beyond the case of an ideal fold, and consider microlensing
by random distributions of point lenses.  This requires numerical
computation of lightcurves, for which we use the microlensing code by
\cite{1999A&A...346L...5W}.

The code is based on the hierarchical tree technique to calculate the
gravitational effect of the lensing masses. The underlying idea is
that when the thin lens approximation is valid, the a microlensing
distribution can be reduced to three planes, the observer plane, the
lens plane and the source plane. The actual effect, namely the
distortion of light rays emitted from a distant source, on the path to
the observer, is numerically computed by through back-tracing:
lightrays are shot backward from the observer to lens plane, where
they become deflected according to the angle dependent on the 2D mass
distribution in the lens plane.  The choice for reversing the photon
motion is motivated by computational resources considerations.  For
the computation of the individual deflection angles the tree concept
is used. The positions of all lensing masses are sorted into a
2-dimensional grid (in the lens plane). Which is subdivided into 4
smaller squares recursively until every cell contains only one
mass. For the actual calculations of the deflection angle not every
mass gravitational influence is treated equally. Instead it is made
use of the fact that the influence of lenses on the considered light
ray falls off with $r^{-1}$ from the point where the ray hits the lens
plane. Hence masses situated at higher distances, can be grouped
together in the calculation of their gravitational potential, using
multipole moments. With this approach the amount of computation time
is reduced. \\

%I don't think this part is necessary

When the deflection angles are determined, the deflected lightrays are
traced to the source plane and a magnification map is created by
counting the number of rays which hit each of its pixels. This is the
inverse of what would happen in a real observation: here an observer
on earth would detect light emitted by several sources with her/his
telescope. Each of these sources emits lightrays which are subject to
the deflection by the intervening gravitational potential of the
lensing mass. However only a fraction of those will end in the
observers telescope on earth. Interesting for the analysis are hence
those lightrays which are originating by the source of interest and
further the subset of these which actually converge at the location of
the observer. So instead of calculating the trajectories of lightrays
emitted by the source in all directions and then succeedingly pick out
the ones which end in the telescope of the observer, the computational
effort can be restricted to these lightrays of interest in the first
place. This is done by treating the problem from the opposite
direction, namely by using the observers location as starting point
and then shooting the lightrays backwards till they hit the source
plane.
% up to here (would remove it)

The result is a pixel map of the number of lightrays which arrive at
the source plane. This intermediate result is denoted as the
magnification map. It characterizes the information about the mass
distribution in the lens plane and its effect on lightrays, and
projects this information onto the source plane. \\\\ Once the map is
created, the lightcurve can be obtained by specifying the transition
path of the source within the map. The code thereby convolves the
function describing the shape of the source with the magnification
pattern of the map for each time step (see equation \ref{eqn:ft2d}).

%this is problematic
 Note that in most observational cases in microlensing, both lens and
 source are moving and further that the lens configuration, and with
 it the magnification pattern, is changing with time. While the first
 subtlety is taken care of by a coordinate transformation in this
 analysis, for the second one the lens configuration is assumed to be
 constant in time.  \\\\ For the model a it was desirable to work with
 a caustic geometry as simple as possible.
% up to here   (would remove it)

When generating the caustic map used in our analysis presented in the
second part of the article only two equal point like masses were used
as lenses, to produce the magnification map depicted in figure
\ref{fig:magnification_map}. The magnification maps calculated by the
Microlensing code are in units of Einstein radii, hence they are scale
free. Physical scale is introduced to them by multiplying the pixels
by the correct distance factor. The Einstein angle is given by
\begin{equation}
\theta_{E}=\sqrt{\frac{4GM}{c^{2}} \frac{D_{LS}}{D_{S}D_{L}}} 
\end{equation}  
and therefore depends on the four parameters
$M,D_{S},D_{L},D_{LS}$. For the cosmological application or quasar
lensing $D_{L}$ is of order 100 Mps to Gpc and $D_{S}$ of order
Gpc.\\ The scale of the magnification map can be converted into a
physical length by multiplying it by a factor of
$\sqrt{\frac{D_{LS}}{D_{S}D_{L}}}$.\\\\ For the next step, the
computation of the actual lightcurve, the code was modified to also
allow for crescent shaped images specified through the parameter set
$R_p,R_n,a,b$. Here $R_p$ denotes the outer radius of the crescent and
$R_n$ the inner one. The orientation of the source image with respect
to the magnification map is specified by the parameters $a$ and $b$ as
the shift of the center of the inner disk from the center of the outer
disk in $x$- and $y$-direction respectively. In the original version
of the code, gaussian and disk shaped images where already
implemented. Those are completely characterized by the single
parameter $R_p$. The values of the parameters are specified in pixel
units corresponding to the magnification map. Further one needs to
specify the start and end point coordinates of the path, which the
center of the source image follows through the magnification map (see
the depiction in figure \ref{fig:magnification_map}). Hence the points
along this path are specified through the number of timesteps for
which the computation of the brightness is to be carried out. Those
points correspond to the actual measurement of the brightness of an
object in the observational case. For each timestep the
two-dimensional convolution of equation \ref{eqn:ft2d} is carried out
numerically for the position of the source on the magnification
map. \\ For the purpose of this analysis it was desirable to mimic the
analytical behaviour of a simple fold as much as possible, for
comparing the numerical result with the analytical one, therefore the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.

>>>>

\section{Numerical microlensing magnification maps}

Having used to simple model of an ideal fold to gain insight, we now
move to the numerical computation of microlensing light curves with
realistic mass distributions.  For this we we used the microlensing
code by \cite{1999A&A...346L...5W}.

The code uses a ray-shooting technique to compute the gravitational
lensing effect of a mass distribution consisting of (a)~a smooth
component and (b)~a random distribution of point masses representing
stars.  The ray shooting maps a grid of $\vec\theta$ to $\vec\beta$
using the lens equation (\ref{eqn:lens}).  That is, the rays are shot
from the observer back to the source.

For the computation of the individual deflection angles
$\vec\alpha(\vec\theta)$ a heirarchical or tree method is used.  The
positions of all lensing masses are put into a grid of $\vec\theta$.
Each grid cell is subdivided into four smaller squares recursively
until every cell contains only one mass.  Nearby masses are added
individually, while distance masses are clumped into larger grid cells
whose net contribution is approximated by its first few multipole
moments.  Scaled units are used, with the constant pre-factor in the
deflection angle (\ref{eq:alpha}) separated out.

The result of ray shooting is a pixel map on the $\vec\beta$ plane of
the number of lightrays which arrive at the source plane from a
particular observer.  This intermediate result is effectively a
magnification map on the source plane.  Once the map is created, the
lightcurve can be obtained by specifying the transit path of the
source across the map.  At each point on this transit line, the code
computes equation (\ref{eqn:ft2d}) convolving the brightness
distribution of the source with the magnification pattern of the map.
In real life, not only are both lens and source moving, but the lens
lens configuration, and with it the magnification pattern, is also
changing with time.  While the first subtlety is taken care of by a
coordinate transformation in this analysis, for the second one the
lens configuration is assumed to be constant in time.

When generating the magnification map depicted in
Figure~\ref{fig:magnification_map}, which is used in our analysis
below, only two point masses were included.  This was done in order to
have clean fold caustics.  For the computation of the actual
lightcurve, the code was modified to also allow for crescent shaped
images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$
denotes the outer radius of the crescent and $R_n$ the inner one. The
orientation of the source image with respect to the magnification map
is specified by the parameters $a$ and $b$ as the shift of the center
of the inner disk from the center of the outer disk in $x$- and
$y$-direction respectively. In the original version of the code,
gaussian and disk shaped images where already implemented. Those are
completely characterized by the single parameter $R_p$. The values of
the parameters are specified in pixel units corresponding to the
magnification map. Further one needs to specify the start and end
point coordinates of the path, which the center of the source image
follows through the magnification map (see the depiction in figure
\ref{fig:magnification_map}). Hence the points along this path are
specified through the number of timesteps for which the computation of
the brightness is to be carried out. Those points correspond to the
actual measurement of the brightness of an object in the observational
case. For each timestep the two-dimensional convolution of equation
\ref{eqn:ft2d} is carried out numerically for the position of the
source on the magnification map.

For the purpose of this analysis it was desirable to mimic the
analytical behaviour of a simple fold as much as possible, for
comparing the numerical result with the analytical one, therefore the
path of the source was chosen, so that it intersects the border of the
caustic perpendicularly, and on a point where the border is a fold
caustic.

### 2015-07-12 paper1/fitting.tex

\begin{figure}
\centering
\includegraphics[width=0.9\hsize,bb=0 0 576 432
                ]{plots/data_cc.eps}
\includegraphics[width=0.9\hsize,bb=0 0 576 432
                ]{plots/data_dd.eps}
\includegraphics[width=0.9\hsize,bb=0 0 576 432
                ]{plots/data_gg.eps}
\caption{\label{fig:mockdata} Three different mock datasets with
  three different parameterisations fitted to each --- crescent (red),
  uniform-disk (blue) and Gaussian disk (green).}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize,bb=0 0 576 432
                  ]{plots/Rp4all.eps}
\caption{\label{fig:mcmc} Posterior probability distribution for the
  source size $R_p$.  The vertical line is the correct value. Same
  color represeents the same dataset whereas same line style
  corresponds to same model fit.  The legend gives the reduced
  $\chi^2$ of the best fit in each case.}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics[width=0.9\hsize,bb=0 0 576 432
                ]{plots/data_cc.eps}
\includegraphics[width=0.9\hsize,bb=0 0 576 432
                ]{plots/data_dd.eps}
\includegraphics[width=0.9\hsize,bb=0 0 576 432
                ]{plots/data_gg.eps}
\caption{\label{fig:mockdata} Three different mock datasets with
  three different parametrisations fitted to each --- crescent (red),
  uniform-disk (blue) and Gaussian disk (green).}
\end{figure}

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize,bb=0 0 576 432
                  ]{plots/Rp4all.eps}
\caption{\label{fig:mcmc} Posterior probability distribution for the
  source size $R_p$.  The vertical line is the correct value. Same
  color represents the same dataset whereas same line style
  corresponds to same model fit.  The legend gives the reduced
  $\chi^2$ of the best fit in each case.}
\end{figure}
